---
title: "Tutorial: Interventions"
author: "Day 4 | Network Modeling for Epidemics"
output:
  html_document:
    toc: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
library(EpiModel)
library(knitr)
knitr::opts_chunk$set(comment = NA)
par(mar = c(3,3,1,1), mgp = c(2,1,0))
```

EpiModel built-in network models also include parameters for a simple intervention that affects the probability of infection given contact between a susceptible and an infected person. This could be, for example, a vaccine that reduces the susceptibility or a condom provision program. This mini-tutorial will show how to implement such an intervention. 

Interventions in this framework have three salient features:

1. They apply to everyone in the population. There is no heterogeneity in who gets the intervention. 
2. The interventions have an efficacy, `inter.eff`, where the invention results in a relative reduction in the probability of infection specified in `inf.prob` by `1-inter.eff`.
3. Interventions start at a defined time step, `inter.start`.

To start, we estimate a very simple temporal ERGM. 
```{r netest, results = "hide", message=FALSE}
nw <- network.initialize(n = 100, directed = FALSE)
formation <- ~edges
target.stats <- 50
coef.diss <- dissolution_coefs(dissolution = ~offset(edges), duration = 20)
est <- netest(nw, formation, target.stats, coef.diss, verbose = FALSE)
```

This model simulates an SI disease in a closed population in which the intervention (e.g., a vaccine) has very strong efficacy for the entire population. The intervention starts at week 25.
```{r siSim, results = "hide"}
param <- param.net(inf.prob = 0.5, inter.eff = 0.96, inter.start = 25)
init <- init.net(i.num = 5)
control <- control.net(type = "SI", nsteps = 100, nsims = 10)
sim <- netsim(est, param, init, control)
plot(sim)
```

This model simulates an SIS disease in which the intervention is slightly less effective (e.g., functional effectiveness of condoms) that starts at week 100.
```{r sisSim, results = "hide"}
param <- param.net(inf.prob = 0.5, inter.eff = 0.8, inter.start = 100, 
                   rec.rate = 0.07)
init <- init.net(i.num = 10)
control <- control.net(type = "SIS", nsteps = 250, nsims = 10)
sim <- netsim(est, param, init, control)
plot(sim)
```


<br>
<hr style="background-color:#909090;height:1px;width:100%">
<small>
*Last updated:* `r format(Sys.time(), "%Y-%m-%d")`
`r if ("EpiModel" %in% names(sessionInfo()$otherPkgs)) paste0(" with EpiModel v", packageVersion("EpiModel"))`
</small>
