---
title: "Tutorial: Dynamic Network Visualization"
author: "Day 4 | Network Modeling for Epidemics"
output:
  html_document:
    toc: no
editor_options: 
  chunk_output_type: console
---

This tutorial shows two forms of dynamic visualization of the spread of infectious disease over networks using the `ndtv` package for R. `ndtv` is part of the larger Statnet suite of software for the representation, modeling, and analysis of network data. It has been designed to work with EpiModel data. 

# Setup

We start by loading both EpiModel and ndtv.

```{r setup, message = FALSE, warning=FALSE}
library("EpiModel")
library("ndtv")
```

In general, the purpose of dynamic network visualization is to gain insight into how the structure of networks over time can influence transmission dynamics. This is related to, but different from, estimating large-scale epidemic properties such as disease incidence or prevalence. Whereas the latter may require large networks (thousands of nodes) over large numbers of time steps (decades of time), visualization usually benefits from a small network size over a short number of time steps. 


# Network Model
This shows an SI epidemic in a closed population where the infection probability is 1, but using an interesting network model that incorporates heterogeneous activity and age mixing (we'll pretend that age is fixed here). That will allow us to see a visualization of the *forward reachable path* where  infection is a deterministic function of the presence of an edge. 

To get started, initialize an empty network and randomly assign demographic attributes for race and age. You may change the parameters for the distributions of these traits.
```{r, eval=FALSE}
nw <- network.initialize(n = 100, directed = FALSE)
nw <- set.vertex.attribute(nw, "race", rbinom(100, 1, 0.5))
nw <- set.vertex.attribute(nw, "age", sample(18:50, 100, TRUE))
```

In this network model, we will model the overall network density, a higher mean degree in the Race 1 group, assortative mixing by age, and a higher than expected propensity for concurrency. 
```{r, eval=FALSE}
formation <- ~edges + nodefactor("race") + absdiff("age") + concurrent
target.stats <- c(40, 30, 20, 80, 20)
coef.diss <- dissolution_coefs(dissolution = ~offset(edges), duration = 20)
est <- netest(nw, formation, target.stats, coef.diss, verbose = FALSE)
```

Since this will be forward reachable path analysis, the transmission probability will equal 1. To demonstrate this visualization, we only need to simulate the epidemic once over a short number of time steps.
```{r, eval=FALSE}
param <- param.net(inf.prob = 1)
init <- init.net(i.num = 1)
control <- control.net(type = "SI", nsteps = 25, nsims = 1, verbose = FALSE)
sim <- netsim(est, param, init, control)
```


# Network Visualization
`ndtv` is an extension package in `Statnet` that allows for dynamic visualization for networks over time. Because this involves animation processing, this may requires some heavy duty computation. 

## Extract and Process the Networks
First we need to extract the network objects from the larger `netsim` object.
```{r getNet, eval = FALSE}
nw <- get_network(sim)
```

Next, we need to add a time-varying nodal attribute to each network that we will animate. This will allow `ndtv` to color the nodes by disease status over time. See the help documentation for `color_tea` to see some of the options and details for this step.
```{r colorTea, eval = FALSE}
nw <- color_tea(nw, verbose = FALSE)
```


## Set Animation Options
There are lots of possibilities for how to model our networks over time. Here we animate the first 25 time steps of the simulation.
```{r aniOptions, eval = FALSE}
slice.par <- list(start = 1, end = 25, interval = 1, 
                  aggregate.dur = 1, rule = "any")
render.par <- list(tween.frames = 10, show.time = FALSE)
plot.par <- list(mar = c(0, 0, 0, 0))
```


## Compute the Dynamic Nodal Coordinates
This step figures out where the nodes should be displayed in each animation frame. There are several dynamic layout options available, but we use the defaults here.
```{r computeAni, eval = FALSE}
compute.animation(nw, slice.par = slice.par, verbose = TRUE)
```

# Create Nodal Attribute Markers
In the network plot, we will use different shapes to distinguish the different race groups, and then use different sizes to show age. Shapes network plots are parameterized by the number of vertex sides, so 4 = a square and 50 = a circle (effectively). For age, we need to divide age by a constant to get the proportions for the sizes in a workable range (this is with trial and error).
```{r, eval=FALSE}
race <- get.vertex.attribute(nw, "race")
race.shape <- ifelse(race == 1, 4, 50)

age <- get.vertex.attribute(nw, "age")
age.size <- age/25
```

## Animate and Save
Finally, we render the animation and save it out to an HTML file using the d3 Javascript library. Note where we have input our nodal attribute markers. This makes for nice-looking animations within webpages. See the `ndtv` help for all the potential options in this step.
```{r renderD3, eval = FALSE}
render.d3movie(
    nw,
    render.par = render.par,
    plot.par = plot.par,
    vertex.cex = age.size,
    vertex.sides = race.shape,
    vertex.col = "ndtvcol",
    edge.col = "darkgrey",
    vertex.border = "lightgrey",
    displaylabels = FALSE,
    filename = paste0(getwd(), "/movie.html"))
```


<br>
<hr style="background-color:#909090;height:1px;width:100%">
<small>
*Last updated:* `r format(Sys.time(), "%Y-%m-%d")`
`r if ("EpiModel" %in% names(sessionInfo()$otherPkgs)) paste0(" with EpiModel v", packageVersion("EpiModel"))`
</small>
