---
title: "Fine-tuning network model estimation EpiModel with set.control.tergm and set.control.ergm"
author: "EpiModel v`r packageVersion('EpiModel')`"
date: "`r Sys.Date()`"
vignette: "%\\VignetteIndexEntry{Working with Model Parameters} %\\VignetteEncoding{UTF-8}
  %\\VignetteEngine{knitr::rmarkdown}\n"
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, message=FALSE}
library(EpiModel)
```

## Introduction

This vignette assumes familiarity with setting up and running network models in EpiModel. See the [Network Modeling for Epidemics](https://epimodel.github.io/sismid/) (NME) course materials and the [EpiModel Gallery](https://epimodel.github.io/EpiModel-Gallery/) for background.

This vignette demonstrates how to:

## Setup

```{r setup, results = "hide"}
rm(list=ls())
set.seed(1)
```

We set up our network:

```{r}
nw <- network_initialize(n = 1000)
nw <- set_vertex_attribute(nw, "age", sample(18:65, 1000, replace = TRUE))
```

## Example 1: approximation works well, no STERGM needed

```{r}
mean_reldur_1 <- 250
coef_diss_1 <- dissolution_coefs(~offset(edges), mean_reldur_1, 0)
formation_1 <- ~edges + degree(0)
target_stats_1 <- c(400, 300)
```

```{r, cache = TRUE}
est_1 <- netest(nw, formation = formation_1, target.stats = target_stats_1, 
                coef.diss = coef_diss_1,  verbose = FALSE)
netdx_1 <- netdx(est_1, nsims = 10, nsteps = 1000) 
plot(netdx_1, mean.smooth = FALSE, qnts.smooth = FALSE)
netdx_1
```

Just check static anyway:

```{r, cache = TRUE}
netdx_stat_1 <- netdx(est_1, dynamic = FALSE, nsims = 1e4) 
plot(netdx_stat_1, mean.smooth = FALSE)
```

# Example 2: approximation is not good, STERGM works

```{r}
mean_reldur_2 <- 10
coef_diss_2 <- dissolution_coefs(~offset(edges), mean_reldur_2, 0)
formation_2 <- ~edges + degree(0)
target_stats_2 <- c(400, 300)
```

```{r, cache = TRUE}
est_2 <- netest(nw, formation = formation_2, target.stats = target_stats_2, 
                coef.diss = coef_diss_2,  verbose = FALSE)
netdx_2 <- netdx(est_2, nsims = 10, nsteps = 1000) 
plot(netdx_2, mean.smooth = FALSE, qnts.smooth = FALSE)
```

Very bad. How are the statics?

```{r, cache = TRUE}
netdx_stat_2 <- netdx(est_2, dynamic = FALSE, nsims = 1e4) 
plot(netdx_stat_2, mean.smooth = FALSE)
```

Very good. So we need to do full STERGM:

```{r, cache = TRUE}
est_2f <- netest(nw, formation = formation_2, target.stats = target_stats_2, 
                coef.diss = coef_diss_2,  verbose = FALSE, edapprox = FALSE)
netdx_2f <- netdx(est_2f, nsims = 10, nsteps = 1000) 
plot(netdx_2f, mean.smooth = FALSE, qnts.smooth = FALSE)
```

That works as is, so no need for further tuning of the estimation algorithm.

## STERGM estimation fails

```{r, cache = TRUE}

formation_3 <- ~edges + degree(0) + nodecov("age")

netdx_2f2 <- netdx(est_2f, nsims = 10, nsteps = 1000, 
                   nwstats.formula = formation_3)

mean_reldur_3 <- 10
coef_diss_3 <- dissolution_coefs(~offset(edges), mean_reldur_3, 0)
target_stats_3 <- c(400, 300, 15000)

#est_3 <- netest(nw, formation = formation_3, target.stats = target_stats_3, 
#                coef.diss = coef_diss_3,  verbose = FALSE, edapprox = FALSE)
```

`Starting Monte Carlo maximum likelihood estimation (MCMLE):`\
`Iteration 1 of at most 60`\
`Sampling`

`Error in ergm.MCMLE(init, s, s.obs, control = control, verbose = verbose,  : `\
`  MCMLE estimation stuck. There may be excessive correlation between model terms, suggesting a poor`\ `model for the observed data. If target.stats are specified, try increasing SAN parameters.`

```{r}
#netdx_stat_3 <- netdx(est_3, dynamic = FALSE, nsims = 1e4) 
#plot(netdx_stat_3, mean.smooth = FALSE)
```
