---
title: 'EpiModel: Mathematical Modeling of Infectious Disease'
author: "Samuel M. Jenness, Steven M. Goodreau, Martina Morris"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EpiModel Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo=FALSE}
vers <- packageVersion("EpiModel")
year <- format(Sys.time(), "%Y")
```

<br>

The **EpiModel** package provides tools for building, solving, and plotting mathematical models of infectious disease in `R`. For the stochastic network modeling of epidemics, it depends on the **Statnet** suite of software for the representation and analysis of networks (see http://statnet.org). **EpiModel** currently provides functionality for three classes of epidemic models and three disease types for modeling. Built-in model classes and types may be expanded by the user for new research.

This vignette is a placeholder for the **EpiModel** tutorials, all hosted online and external to the package to minimize overhead and provide flexibility in updates. Everything may be found at the **EpiModel** website: http://epimodel.org/

Within the package, please consult the extensive help documentation:
```{r, eval=FALSE}
help(package = "EpiModel")
```

To see the latest updates to **EpiModel**, consult the software `NEWS`:
```{r, eval=FALSE}
news(package = "EpiModel")
```

If using the **EpiModel** package for research, we request that you acknowledge this by citing the software in any publications:

> Jenness SM, Goodreau SM, Morris M (`r year`). *EpiModel: Mathematical Modeling of Infectious Disease.* R Package Version `r vers`. URL: http://epimodel.org/.
