---
title: "Working with Model Parameters in EpiModel"
author: "EpiModel v`r packageVersion('EpiModel')`"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Working with Model Parameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, message=FALSE}
library(EpiModel)
```


## Introduction

This vignette assumes familiarity with setting up and running network models in EpiModel. See the [Network Modeling for Epidemics](https://epimodel.github.io/sismid/) (NME) course materials and the [EpiModel Gallery](https://epimodel.github.io/EpiModel-Gallery/) for background.

For working with nodal attributes and epidemic summary statistics, see the companion vignette *Working with Custom Attributes and Summary Statistics in EpiModel*. For network objects and edgelists, see *Working with Network Objects in EpiModel*.

In a model, *parameters* are the input variables that define aspects of system behavior. In a basic SIS (Susceptible-Infected-Susceptible) model, these might include an **infection probability**, an **act rate**, and a **recovery rate**. In simple models, each parameter is a single fixed value that remains constant throughout a simulation. Real-world applications, however, often require more flexibility: parameters that change over time (e.g., to represent an intervention rollout or behavioral change during an epidemic), parameters drawn from distributions (for sensitivity analysis or calibration), or large parameter sets managed via spreadsheets.

This vignette demonstrates how to implement:

- **Scenarios:** Named sets of parameter changes applied at the start of a simulation or at specific time steps---the primary mechanism for time-varying parameters, counterfactual analysis, and intervention modeling.
- **Parameter input via tables:** Passing many parameters at once through a `data.frame`, enabling spreadsheet-based parameter management.
- **Parameter uncertainty:** Drawing parameter values from distributions or from joint parameter sets and running each draw as a scenario, for uncertainty and sensitivity analysis.
- **Parameter and control updaters (advanced):** The low-level mechanism underlying scenarios, exposed directly for cases where the scenario API is not flexible enough.

## Scenarios

Scenarios allow you to define named sets of parameter values that can change at specific time steps during a simulation. This is the recommended approach for implementing time-varying parameters---for example, modeling an intervention that reduces transmission probability partway through an epidemic, or comparing multiple policy alternatives.

Our research paper implementing time-varying parameter scenarios for COVID-19 sexual distancing interventions was published in the *Journal of Infectious Diseases*. The code [can be found here](https://github.com/EpiModel/SexualDistancing).

### Setup

First, we set up a simple SIS model as the base case.

```{r setup, results = "hide"}
set.seed(10)

nw <- network_initialize(n = 200)
est <- netest(nw,
  formation = ~edges, target.stats = 60,
  coef.diss = dissolution_coefs(~offset(edges), 10, 0),  # duration = 10, closed population
  verbose = FALSE
)

param <- param.net(inf.prob = 0.9, rec.rate = 0.01, act.rate = 2)
init <- init.net(i.num = 10)
control <- control.net(type = "SIS", nsims = 1, nsteps = 250, verbose = FALSE)
```

### Scenario Definitions

Scenarios are defined using `create_scenario_list`, which takes a specially formatted `data.frame` as input and outputs a list of scenarios usable by EpiModel. We use `tibble::tribble` here for readability, but any `data.frame` constructor works.

```{r scenarios_setup}
library(dplyr)

scenarios.df <- tribble(
  ~.scenario.id, ~.at, ~inf.prob, ~rec.rate,
  "base", 0, 0.9, 0.01,
  "initial_change", 0, 0.2, 0.01,
  "multiple_changes", 0, 0.1, 0.04,
  "multiple_changes", 100, 0.9, 0.01,
  "multiple_changes", 200, 0.1, 0.1
)

knitr::kable(scenarios.df)
```

The `data.frame` requires two columns:

- `.scenario.id`: an identifier for the scenario.
- `.at`: the time step at which the parameter changes should apply. Use `0` for changes that take effect before the simulation begins.

Remaining columns are parameter names. They must start with a letter and contain only letters, numbers, or `.` (underscores are reserved for vector indexing, described below). If a cell is `NA`, EpiModel sets that parameter to `NA`---it does *not* mean "leave unchanged."

In this example, the `"multiple_changes"` scenario has three rows sharing the same `.scenario.id`, representing a single scenario with parameter changes at three different time points.

To convert from the `data.frame` to a usable list of scenarios:

```{r scenarios_list}
scenarios.list <- create_scenario_list(scenarios.df)
str(scenarios.list, max.level = 2)
```

### Running Scenarios

We loop over all scenarios, using `use_scenario` to create a modified parameter object for each. Under the hood, `use_scenario` generates parameter updaters (described in the Advanced section below) that apply the specified changes at the designated time steps.

```{r simulations}
# List to hold simulation results
d_list <- vector(mode = "list", length = length(scenarios.list))
names(d_list) <- names(scenarios.list)

for (scenario in scenarios.list) {
  print(scenario$id)
  sc.param <- use_scenario(param, scenario)
  sim <- netsim(est, sc.param, init, control)
  d_sim <- as.data.frame(sim)
  d_sim[["scenario"]] <- scenario$id
  d_list[[scenario$id]] <- d_sim
}
```

For the `"multiple_changes"` scenario, messages appear at time steps 100 and 200 indicating that parameters were modified during the simulation. The other scenarios are silent because their changes occur before the simulation begins (`.at = 0`).

Now we can plot the results. We use base R plotting here rather than `plot.netsim()` because we need to overlay results from separate `netsim` runs on a single axis.

```{r plotting, fig.width = 8}
plot(d_list$base$time, d_list$base$i.num,
     type = "l", col = 1, lwd = 2, ylim = c(0, 250),
     xlab = "Time Step", ylab = "Number Infected")
lines(d_list$initial_change$time, d_list$initial_change$i.num,
      type = "l", col = 2, lwd = 2)
lines(d_list$multiple_changes$time, d_list$multiple_changes$i.num,
      type = "l", col = 3, lwd = 2)
abline(v = c(100, 200), lty = 2)
legend("topleft", legend = names(d_list),
       col = 1:3, lwd = 2, cex = 0.9, bty = "n")
```

The `"initial_change"` scenario shows a slower epidemic than the base case because the infection probability is reduced from 0.9 to 0.2 from the start.

The `"multiple_changes"` scenario demonstrates three phases:

- Steps 1--99: Slow growth (low infection probability, higher recovery rate).
- Steps 100--199: Base-case parameters applied, so the epidemic trajectory accelerates.
- Steps 200--250: Infection probability drops and recovery rate increases, driving rapid epidemic decline.

### Scenarios with Parameter Vectors

In full-scale modeling projects, parameters are often vectors. For example, `hiv.test.rate` might be a vector of length 3 containing weekly HIV screening probabilities for Black, Hispanic, and White persons.

To use vectors in scenarios, create a separate column for each element using the naming convention `paramname_N`, where `N` is the position in the vector:

```{r multi_param}
scenarios.df <- tribble(
  ~.scenario.id, ~.at, ~hiv.test.rate_1, ~hiv.test.rate_2, ~hiv.test.rate_3,
  "base", 0, 0.001, 0.001, 0.001,
  "initial_change", 0, 0.002, 0.001, 0.002,
  "multiple_changes", 0, 0.002, 0.001, 0.002,
  "multiple_changes", 100, 0.004, 0.001, 0.004,
  "multiple_changes", 200, 0.008, 0.001, 0.008
)
knitr::kable(scenarios.df)
```

All elements of a vector parameter must be specified in the `data.frame`, even if some values do not change across scenarios (e.g., `hiv.test.rate_2` is constant here but must still be included).

When working with many parameters, we recommend storing the `scenarios.df` as a CSV, RDS or Excel file for easier sharing and editing.

```{r, eval = FALSE}
# CSV
write.csv(scenarios.df, "scenarios.csv", row.names = FALSE)
scenarios.df <- read.csv("scenarios.csv")

# RDS
saveRDS(scenarios.df, "scenarios.rds")
scenarios.df <- readRDS("scenarios.rds")
```

## Parameter Input via Table

The `param.net` function accepts a `data.frame` of parameters through the `data.frame.params` argument. This allows passing many parameters at once or working with a spreadsheet to track and update parameters before use in EpiModel.

The `data.frame` requires three columns:

1. `param`: The parameter name. For vector parameters (length > 1), append the position with an underscore (e.g., `"p_1"`, `"p_2"`).
2. `value`: The parameter value (as a character string).
3. `type`: The type of the parameter value. Only three values are accepted: `"numeric"`, `"logical"`, or `"character"`.

Additional columns (e.g., `details`, `source`) may be included for documentation but are ignored by EpiModel.

```{r param_table}
df_params <- tribble(
  ~param, ~value, ~type,
  "hiv.test.rate_1",  "0.003",        "numeric",
  "hiv.test.rate_2",  "0.102",        "numeric",
  "hiv.test.rate_3",  "0.492",        "numeric",
  "prep.require.lnt", "TRUE",         "logical",
  "group_1",          "first",        "character",
  "group_2",          "second",       "character"
)
knitr::kable(df_params)
```

Pass the table via `data.frame.params`. These parameters can be combined with named parameters for maximum flexibility. In case of conflict, named parameters take priority over those in the `data.frame`:

```{r param_table2}
param <- param.net(data.frame.params = df_params,
                   other.param = c(5, 10), act.rate = 1)
param
```


## Parameter Uncertainty

Fixed parameters assume that each input value is known with certainty. In practice, many inputs are estimated with error, and an analysis may need to show how that uncertainty propagates to the epidemic outcomes. In EpiModel this is done with scenarios: the parameter values are drawn before any simulation runs and stored in a `data.frame` with one row per draw, and each row is run as a scenario that replaces the base parameter values at `.at = 0`. Because the draws exist as a table before the simulations start, they can be inspected and saved, they can be joined back onto the simulation output, and parameters that were sampled together stay together.

### Drawing Parameter Values

We draw eight sets of values for the act rate and the per-act infection probability of the SIS model from the setup section. Each parameter is drawn independently with base R sampling functions: `sample()` selects from a discrete set of values (with unequal weights given through its `prob` argument), and `rbeta()` draws from a continuous distribution.

```{r uncertainty_draws}
set.seed(12)
n.draws <- 8

draws <- data.frame(
  .scenario.id = paste0("draw", seq_len(n.draws)),
  .at = 0,
  act.rate = sample(c(1, 2, 3), n.draws, replace = TRUE),
  inf.prob = rbeta(n.draws, 2, 6)
)
knitr::kable(draws, digits = 3)
```

The `.scenario.id` and `.at` columns follow the scenario format described above. Each draw is its own scenario, and the remaining columns hold the drawn values.

### Correlated Draws

Independent draws are appropriate only when the parameters are independent. Parameter sets from a calibration, posterior samples, and sampling designs with an imposed correlation structure must instead be drawn as whole rows. Drawing each column separately from such a table would discard the correlation between parameters.

As an example, suppose that a calibration produced 500 accepted parameter sets in which the act rate and the infection probability are negatively correlated, as happens when the calibration targets inform their combined effect on transmission but not each parameter separately.

```{r uncertainty_posterior}
posterior <- data.frame(act.rate = runif(500, 1, 3))
posterior$inf.prob <- 0.5 / posterior$act.rate * exp(rnorm(500, 0, 0.1))
cor(posterior$act.rate, posterior$inf.prob)
```

To keep each set together, we sample rows of the table rather than values from each column:

```{r uncertainty_joint}
draws.joint <- data.frame(
  .scenario.id = paste0("draw", seq_len(n.draws)),
  .at = 0,
  posterior[sample(nrow(posterior), n.draws), ],
  row.names = NULL
)
knitr::kable(draws.joint, digits = 3)
```

This table is run in the same way as the independent draws below. If every accepted parameter set can be simulated, the rows of `posterior` can be used directly without sampling.

### Vector Parameters

Vector parameters use the `paramname_N` columns described in *Scenarios with Parameter Vectors*, with one column per element. Each element can be drawn from its own distribution, as in this example of a testing rate stratified into two groups (which a custom testing module would use), or the elements can be drawn jointly as rows. As with any scenario, the parameter must already be defined in the base parameter object.

```{r uncertainty_vector}
draws.vec <- data.frame(
  .scenario.id = paste0("draw", 1:3),
  .at = 0,
  test.rate_1 = rnorm(3, 0.05, 0.01),
  test.rate_2 = rnorm(3, 0.15, 0.03)
)

param.vec <- param.net(inf.prob = 0.3, test.rate = c(0.05, 0.15))
sc.vec <- create_scenario_list(draws.vec)
use_scenario(param.vec, sc.vec$draw2)$test.rate
```

### Running the Draws

We convert the independent draws to a scenario list and run each one as in *Running Scenarios* above, with two simulations per draw.

```{r uncertainty_run}
param <- param.net(inf.prob = 0.3, act.rate = 2, rec.rate = 0.05)
control <- control.net(type = "SIS", nsims = 2, nsteps = 100, verbose = FALSE)

draws.list <- create_scenario_list(draws)

d_list <- lapply(draws.list, function(scenario) {
  sc.param <- use_scenario(param, scenario)
  sim <- netsim(est, sc.param, init, control)
  d_sim <- as.data.frame(sim)
  d_sim[["scenario"]] <- scenario$id
  d_sim
})
d_unc <- bind_rows(d_list)
```

The values of `inf.prob` and `act.rate` in `param` are placeholders, since every scenario replaces them before its simulations start. The parameters a simulation actually used are stored in its output: `sim$param` holds the base parameters with the scenario values applied, along with the scenario ID in `.scenario.id`.

The two simulations within a draw differ only by stochastic variation, while differences between draws also reflect the parameter uncertainty. Setting `nsims = 1` with a larger number of draws gives one simulation per draw. For large numbers of draws, the [EpiModelHPC](https://github.com/EpiModel/EpiModelHPC) package runs a scenario list locally with `netsim_scenarios()` or as an array job on a SLURM cluster with `step_tmpl_netsim_scenarios()`.

### Linking Draws to Outcomes

The scenario ID links each simulation to its row of the draws table, so the drawn values can be joined onto the epidemic output. Here we also compute the probability of transmission within a discordant partnership per time step, `1 - (1 - inf.prob)^act.rate`, and compare it with prevalence at the final time step.

```{r uncertainty_join}
d_unc <- left_join(d_unc, select(draws, -.at), by = c("scenario" = ".scenario.id"))
d_unc$trans.prob <- 1 - (1 - d_unc$inf.prob)^d_unc$act.rate

d_final <- filter(d_unc, time == max(time))
```

```{r uncertainty_plot, fig.width = 8}
plot(d_final$trans.prob, d_final$i.num / d_final$num,
     pch = 19, col = d_final$act.rate, ylim = c(0, 1),
     xlab = "Transmission probability per partnership per time step",
     ylab = "Prevalence at time step 100")
legend("topleft", legend = paste("act.rate =", 1:3),
       col = 1:3, pch = 19, bty = "n")
```

Each point is one simulation. In this model, prevalence at the final time step increases with the per-partnership transmission probability, and the two simulations for each draw show the stochastic variation around that relationship.

### Migrating from `random.params`

Through EpiModel v2.6.2, random parameters were specified with the `random.params` argument of `param.net()`, and `netsim()` drew one set of values at the start of each simulation. That interface has been removed, and passing `random.params` to `param.net()` now produces an error. Each part of it has a counterpart in the approach above, where `n` is the number of draws:

| Removed interface | Replacement |
|:--|:--|
| `param_random(values, prob)` | `sample(values, n, replace = TRUE, prob = prob)` |
| A generator function such as `function() rbeta(1, 1, 2)` | The vectorized draw `rbeta(n, 1, 2)` |
| A generator function returning a vector | One `paramname_N` column per element |
| `param.random.set` | Sampling whole rows of the table, as in *Correlated Draws* |
| `get_param_set(sim)` | The table of draws, joined to the output by scenario ID |


## Time-Varying Parameters and Control Settings (Advanced)

The *scenarios* system described above uses an internal updater module to implement parameter changes at designated time steps. This section describes the underlying updater mechanism directly. Use this when the scenario API is not flexible enough---for example, when you need relative (function-based) parameter changes or time-varying *control* settings.

### Parameter Updaters

An updater is a `list` with two required elements: `at` (the time step for the change) and `param` (a named list of new parameter values):

```{r updater-example}
list(
  at = 10,
  param = list(
    inf.prob = 0.3,
    act.rate = 0.5
  )
)
```

This updater sets `inf.prob` to 0.3 and `act.rate` to 0.5 at time step 10. Multiple updaters are combined in a list:

```{r updater-list}
list.of.updaters <- list(
  list(
    at = 100,
    param = list(
      inf.prob = 0.3,
      act.rate = 0.3
    )
  ),
  list(
    at = 125,
    param = list(
      inf.prob = 0.01
    )
  )
)
```

### Incorporating Updaters

Pass the updater list to `param.net` via the `.param.updater.list` argument:

```{r updater-module}
param <- param.net(
 inf.prob = 0.1,
 act.rate = 0.1,
 .param.updater.list = list.of.updaters
)
init <- init.net(i.num = 10)
control <- control.net(
 type = "SI",
 nsims = 1,
 nsteps = 200,
 verbose = FALSE
)
```

```{r updater-module2}
nw <- network_initialize(n = 100)
est <- netest(
  nw,
  formation = ~edges,
  target.stats = 50,
  coef.diss = dissolution_coefs(~offset(edges), 10, 0),  # duration = 10, closed population
  verbose = FALSE
)
mod <- netsim(est, param, init, control)
```

The plot shows inflection points at the two parameter change points. At step 100, both `inf.prob` and `act.rate` increase from 0.1 to 0.3, accelerating transmission. At step 125, `inf.prob` drops to 0.01, slowing the epidemic.

```{r}
plot(mod, mean.smooth = FALSE)
abline(v = c(100, 125), lty = 2, col = "grey50")
```


#### Verbosity

Each updater can include an optional `verbose` element. When `TRUE` (the default), EpiModel prints a message describing the changes:

```{r updater-verbose}
list(
  at = 10,
  param = list(
    inf.prob = 0.3,
    act.rate = 0.5
  ),
  verbose = TRUE
)
```

#### Relative Parameter Changes

Instead of setting a parameter to a fixed new value, you can define a *function* that transforms the current value. This is useful for multiplicative changes or logit-scale adjustments.

```{r updater-function}
list(
  at = 10,
  param = list(
    inf.prob = function(x) plogis(qlogis(x) + log(2)),
    act.rate = 0.5
  )
)
```

In this updater, `act.rate` is set to the fixed value 0.5 as before. For `inf.prob`, instead of a value, we provide a function. At time step 10, EpiModel applies this function to the *current* value of `inf.prob`. If `inf.prob` was set to 0.1 by `param.net`, the function computes `plogis(qlogis(0.1) + log(2))` = `r round(plogis(qlogis(0.1) + log(2)), 4)`, which doubles the odds of transmission while keeping the result on the probability scale.

### Time-Varying Control Settings

Control settings can also be changed during a simulation using the same updater mechanism. Each control updater uses a `control` element (instead of `param`):

```{r control-updater-example}
list.of.updaters <- list(
  list(
    at = 100,
    control = list(
      resimulate.network = FALSE
    )
  ),
  list(
    at = 125,
    control = list(
      verbose = FALSE
    )
  )
)
```

This example turns off network resimulation at step 100 and disables model verbosity at step 125.

Pass control updaters to `control.net` via `.control.updater.list`, paralleling how parameter updaters are passed to `param.net` via `.param.updater.list`:

```{r}
control <- control.net(
 type = "SI",
 nsims = 1,
 nsteps = 200,
 verbose = TRUE,
 .control.updater.list = list.of.updaters
)
```
