---
title: "Working With Model Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model-parameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, include = FALSE}
library(EpiModel)
```

## Introduction

This vignette discusses mechanisms usable inside `EpiModel` network models with custom extension modules. More information about these may be found in the "New Network Models with EpiModel" section of the [EpiModel Tutorials](http://www.epimodel.org/tut.html).

In a model, *parameters* are the input variables used to define aspects of the system behavior. In the basic built-in SIS (Susceptible-Infected-Susceptible) model, these parameters could be the **act rate**, the **infection probability** and the **recovery rate**. In simple models, each of these parameters are single fixed values that do not change over the course of a simulation. In more complex models, we may want more flexibility in model parameterization.

Therefore, in this vignette, we demonstrate how to implement:

- Random parameters, which have a distribution of possible values rather than a single fixed value.
- Time-varying parameters, which may change at specific time steps during the simulation.

## Random Parameters

### The Model

First, we define a simple SI model.

```{r model}
nw <- network_initialize(n = 50)

est <- netest(
  nw, formation = ~edges,
  target.stats = 25,
  coef.diss = dissolution_coefs(~offset(edges), 10, 0),
  verbose = FALSE
)

param <- param.net(
  inf.prob = 0.3,
  act.rate = 0.5,
  dummy.param = 4,
  dummy.strat.param = c(0, 1)
)

init <- init.net(i.num = 10)
control <- control.net(type = "SI", nsims = 1, nsteps = 5, verbose = FALSE)
mod <- netsim(est, param, init, control)
mod
```

In the parameters, we set the value for `inf.prob` and `act.rate` as fixed, but we also define `dummy.param` and `dummy.strat.param`. These latter two parameters will serve to illustrate random parameters. `dummy.strat.param` has two elements; this may be used, for example, to stratify a parameter by subpopulation.

The last line prints a summary of the model. In it we can see the value of the parameters under the "Fixed Parameters" section. Note the additional `groups` parameter defined automatically by `EpiModel` as part of the "SI" model definition.

### Adding Random Parameters

To allow our parameters to be drawn from a distribution of random values, we use the `random.params` argument to `param.net`. There are two ways of defining which parameters are random, and the distribution of values for those random parameters to draw randomly and how to draw them.

#### Generator Functions

The first option is to define a generator function for each parameter we want to treat as random.

```{r generators}
my.randoms <- list(
  act.rate = param_random(c(0.25, 0.5, 0.75)),
  dummy.param = function() rbeta(1, 1, 2),
  dummy.strat.param = function() c(
    rnorm(1, 0.05, 0.01),
    rnorm(1, 0.15, 0.03)
  )
)

param <- param.net(
  inf.prob = 0.3,
  random.params = my.randoms
)

param
```

Here we kept the `inf.prob` parameter fixed at `0.3` and defined a `list` object called `my.randoms` containing 3 elements:

- `act.rate` uses the `param_random` [function factory](https://adv-r.hadley.nz/function-factories.html) defined by EpiModel
  (see `?EpiModel::param_random`).
- `dummy.param` is a function with no argument that returns a random value from a beta distribution.
- `dummy.strat.param` is a function with no argument that returns 2 values sampled from normal distributions, each with different means and standard deviations.

Each element is named after the parameter it will fill and MUST BE a function taking no argument and outputting a vector of the right size for the parameter: size 1 for `act.rate` and `dummy.param`; size 2 for `dummy.strat.param`.

The rest of the model is run as before, although we increase the simulation count to three to demonstrate the parameter stochasticity.
```{r generators_run}
control <- control.net(type = "SI", nsims = 3, nsteps = 5, verbose = FALSE)
mod <- netsim(est, param, init, control)

mod
```

After running 3 simulations we can see that 2 parameters are still displayed under the "Fixed Parameters" section: `inf.prob` and `groups`. The other ones are displayed under the "Random Parameters". `act.rate` and `dummy.param` now have 3 values associated with them, one per simulation. `dummy.strat.param` have `<list>` as value because each simulation has a vector of size 2.

We can inspect the values with the `get_param_set` function:

``` {r generators_inspect}
get_param_set(mod)
```

#### Parameter set

The drawback of generator function approach above is that it cannot produce correlated parameters. For instance, we may want `dummy.param` and `dummy.strat.param` to be related to one another within a single simulation. We might also use another, external approach for generating parameter sets (e.g., Latin hypercube sampling of multiple parameters).

For this, we need to pre-define a `data.frame` of the possible values:

```{r set_df}
n <- 5

related.param <- data.frame(
  dummy.param = rbeta(n, 1, 2)
)

related.param$dummy.strat.param_1 <- related.param$dummy.param + rnorm(n)
related.param$dummy.strat.param_2 <- related.param$dummy.param * 2 + rnorm(n)

related.param
```

We now have a `data.frame` with 5 rows and 3 columns. Each row contains a set of parameters values that will be used together in a model. This way we keep the relationship between each value.

The first column of the `data.frame` is named `dummy.param`, similar to the name of the parameter. For `dummy.start.param` we need two columns as the parameter contains two values. To achieve this, we name the 2 columns `dummy.start.param_1` and
`dummy.strat.param_2`. The value after the underscore informs `EpiModel` how it should combine these values. This in turn means that the underscore symbol is not allowed in proper parameter names.

Then we set up the rest of the parameters. `related.param` is saved in the `my.randoms` list under the special reserved name `param_random_set`.

```{r set_param}
my.randoms <- list(
  act.rate = param_random(c(0.25, 0.5, 0.75)),
  param_random_set = related.param

)

param <- param.net(
  inf.prob = 0.3,
  random.params = my.randoms
)

param
```

Notice here that we combined a generator function for `act.rate` and a set of
correlated parameters with `param_random_set`.

```{r set_run}
control <- control.net(type = "SI", nsims = 3, nsteps = 5, verbose = FALSE)
mod <- netsim(est, param, init, control)

mod
```

The output is similar to what we saw with the generator functions.


## Time-Varying Parameters

When doing modeling work, the researcher may want to simulate interventions or events that vary over time. In (our 2021  paper)[https://academic.oup.com/jid/article/223/6/1019/6122459], we simulated the effects of COVID-19–Related behavior change and clinical service
interruption on HIV and STI incidence. These included simulations with systemic events with a beginning and an end. To work with these situations, EpiModel offers an optional module to programmatically handle parameter changes across time steps.

This vignette assumes a basic understanding of EpiModel custom models. If the code does not make sense, please check the [EpiModel Tutorials](http://www.epimodel.org/tut.html).

### Parameter Updaters

To define what parameters should change and when during the simulation, we need to define **updaters**. An updater is a `list` with to named elements: `at`,
the time step when the change will take place, and `param` a named list of parameters to update.

```{r updater-example, eval = FALSE}
list(
  at = 10,
  param = list(
    inf.prob = 0.3,
    act.rate = 0.5
  )
)
```

This example defines an updater that will change the value of the `inf.prob` parameter to `0.3` and the value to the `act.rate` parameter to `0.5`. This
change will happen during the 10th time step.

As mentioned above, we usually want to define several changes at once. EpiModel accept a `list` of **updaters**.


```{r updater-list}
# Create a `list.of.updaters`
list.of.updaters <- list(
  # this is one updater
  list(
    at = 100,
    param = list(
      inf.prob = 0.3,
      act.rate = 0.3
    )
  ),
  # this is another updater
  list(
    at = 125,
    param = list(
      inf.prob = 0.01
    )
  )
)

 # The `list.of.updaters` goes into `param.net` under `param.updater.list`
 param <- param.net(
   inf.prob = 0.1,
   act.rate = 0.1,
   param.updater.list = list.of.updaters
 )
```

In this example, we define 2 **updaters**, one that occurs at time step 100 and the other one at time step 20. In practice, `inf.prob` and `act.rate` are set to `0.1` at the beginning by `param.net`, at time step 100 they are both updated to `0.3` and at time step 125 `act.rate` is reduced to `0.01`.

### Enabling the `updater` Module

This functionality is part of an optional module provided by `EpiModel`, `updater.net`. This module must then be enabled in `control.net`.

Below we set up a complete example with a closed population SI model using the parameters and **updaters** defined above. We then plot the size of the
infected and susceptible population over time to see the effects of the **updaters**.

```{r updater-module, fig.align = "center", fig.height = 4, fig.width = 6}
 control <- control.net(
   type = NULL, # must be NULL as we use a custom module
   nsims = 1,
   nsteps = 200,
   verbose = FALSE,
   updater.FUN = updater.net,
   infection.FUN = infection.net
 )

nw <- network_initialize(n = 50)
nw <- set_vertex_attribute(nw, "race", rbinom(50, 1, 0.5))
est <- netest(
  nw,
  formation = ~edges,
  target.stats = 25,
  coef.diss = dissolution_coefs(~offset(edges), 10, 0),
  verbose = FALSE
)

init <- init.net(i.num = 10)
mod <- netsim(est, param, init, control)

plot(mod)
```

This is the minimal example to simplify the demonstration of these time-varying parameters.

**Important**: when enabling the `updater.net` module, the researcher must pay particular attention to the order the modules are run. It is recommended to have
the `updater.net` module run first so the changes happen at the beginning of the time step.

Some functionality of `updater.net` were not described above and are explained in `?EpiModel::updater.net`.

#### Verbosity

When creating an **updater**, one can add an optional `verbose` element to the `list`. If `TRUE`, `updater.net` will output a `message` describing what changes
where performed when it occurs during the simulation.

```{r updater-verbose, echo = FALSE}
list(
  at = 10,
  param = list(
    inf.prob = 0.3,
    act.rate = 0.5
  ),
  verbose = TRUE
)
```

#### Relative Parameter Changes

It is sometime useful to configure the changes with respect to the current value instead of a fixed new value. This is possible, as demonstrated below.

```{r updater-function, echo = FALSE}
list(
  at = 10,
  param = list(
    inf.prob = function(x) plogis(qlogis(x) + log(2)),
    act.rate = 0.5
  )
)
```

This updater will set the value of `act.rate` to `0.5` as we saw earlier. But, for `inf.prob` we put a `function` (not a function call). In this case,
`updater.net` will apply the `function` to the current value of `act.rate`. If we consider as in the previous example that `act.rate` is set to `0.1` by
`param.net`, then its new value will be obtained by adding an Odds Ratio of 2 to the original value `plogis(qlogis(0.1) + log(2))`: `0.1818182`.
