---
title: "Updating-the-model-parameters-during-runtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updating-the-model-parameters-during-runtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(EpiModel)
```

# Introduction

When doing modeling work, the researcher may want to simulate interventions or
events that vary over time. In (their 2021
paper)[https://academic.oup.com/jid/article/223/6/1019/6122459] Jenness & al
simulate the effects of COVID-19–Related Sexual Distancing and Clinical Service
Interruption on HIV and STI incidence. These are events with a beginning and an
end. To work with these situations, EpiModel offers an optional module to
programmatically handle the changes in the model over time.

This vignette assumes a basic understanding of EpiModel custom models. If the
code does not make sense, please check the
[EpiModel tutorials](http://www.epimodel.org/tut.html).

# Parameter updaters

To define what parameters should change and when during the simulation, we need
to define **updaters**. An updater is a `list` with to named elements: `at`,
the time step when the change will take place, and `param` a named list of
parameters to update.

```{r updater-example, echo = FALSE}
list(
  at = 10,
  param = list(
    inf.prob = 0.3,
    act.rate = 0.5
  )
)
```

This example defines an updater that will change the value of the `inf.prob`
parameter to `0.3` and the value to the `act.rate` parameter to `0.5`. This
change will happen during the 10th time step.

As mentioned above, we usually want to define several changes at once. EpiModel
accept a `list` of **updaters**.


```{r updater-list, echo = FALSE}
# Create a `list.of.updaters`
list.of.updaters <- list(
  # this is one updater
  list(
    at = 10,
    param = list(
      inf.prob = 0.5,
      act.rate = 0.5
    )
  ),
  # this is another updater
  list(
    at = 20,
    param = list(
      inf.prob = 0.1,
      act.rate = 0.1
    )
  )
)

 # The `list.of.updaters` goes into `param.net` under `param.updater.list`
 param <- param.net(
   inf.prob = 0.1,
   act.rate = 0.1,
   param.updater.list = list.of.updaters
 )
```

In this example, we define 2 **updaters**, one that occurs at time step 10 and
the other one at time step 20. In practice, `inf.prob` and `act.rate` are set to
`0.1` at the beginning by `param.net`, at time step 10 they are both updated to
`0.5` and at time step 20 their values go back to the original `0.1`.

# Enabling the `updater` module

This functionality is part of an optional module provided by `EpiModel`,
`updater.net`. This module has to be enabled in `control.net`.

```{r updater-module, echo = FALSE}
 control <- control.net(
   type = NULL, # must be NULL as we use a custom module
   nsims = 1,
   nsteps = 20,
   verbose = FALSE,
   updater.FUN = updater.net
)
```

This is the minimal `control.net` to enable the module. Usually, a researcher
will want to add it to a more complex model with several bespoke modules. See the
"Advanced Extension Models" section in the
[EpiModel tutorials](http://www.epimodel.org/tut.html).

**Important**: when enabling the `updater.net` module, the researcher must pay
particular attention to the order the modules are run. It is recommended to have
the `updater.net` module run first so the changes happen at the beginning of
the time step.

# Advanced functionalities

Some functionalities of `updater.net` were not described above and are explained
in `?EpiModel::updater.net`.

## Verbosity

When creating an **updater**, one can add an optional `verbose` element to the
`list`. If `TRUE`, `updater.net` will output a `message` describing what changes
where performed when it occurs during the simulation.

```{r updater-verbose, echo = FALSE}
list(
  at = 10,
  param = list(
    inf.prob = 0.3,
    act.rate = 0.5
  ),
  verbose = TRUE
)
```

## Relative parameter change

It is sometime useful to configure the changes with respect to the current value
instead of a fixed new value.

```{r updater-function, echo = FALSE}
list(
  at = 10,
  param = list(
    inf.prob = function(x) plogis(qlogis(x) + log(2)),
    act.rate = 0.5
  )
)
```

This updater will set the value of `act.rate` to `0.5` as we saw earlier. But,
for `inf.prob` we put a `function` (not a function call). In this case,
`updater.net` will apply the `function` to the current value of `act.rate`. If
we consider as in the previous example that `act.rate` is set to `0.1` by
`param.net`, then its new value will be obtained by adding an Odds Ratio of 2
to the original value `plogis(qlogis(0.1) + log(2))`: `0.1818182`.



