---
title: "Exploring degrees of freedom"
author: "NME Team"
output:
  html_document:
    fig_width: 8
    highlight: kate
    theme: cosmo
    toc: yes
    toc_float: true
---

```{r, message=FALSE}
library(EpiModel)
messageCtrl <- "FALSE"
textRslt <- "hide"
```

The code below explore the estimates we get for overspecified
models -- using EpiModel and ergm directly.  Setup is based on Sam's lab
(d3-s4.R 2019).  Models 1 and 2 are identifiable, because they
specify 4 or fewer terms for the degree distribution matrix.  Models 3-5 are overspecified.  Note
that `degrange(from=4)` does not contribute to overspecification.

All models typically fit easily, without warnings, but for the overspecified models:  

* the coefs vary (because there is no unique solution) 
* different coef solutions reproduce the target stats (so each is a valid solution)
* the SE's are nonsensical (e+5) or NA  
* the MCMC% is sometimes 100  
* sometimes the MCMLE estimation hangs  

## Setup
```{r, messages=messageCtrl}
nb1 <- nb2 <- 250
nw <- network.initialize(n = nb1 + nb2,
                         bipartite = nb1,
                         directed = FALSE)
nw

# Degree distributions by mode
deg.frac1 <- c(0.40, 0.55, 0.04, 0.01)
deg.frac2 <- c(0.54, 0.31, 0.10, 0.05)
deg.num1 <- round(deg.frac1*250,0)
deg.num2 <- round(deg.frac2*250,0)
cbind(dist1=deg.num1, dist2=deg.num2)

# Given groupsizes and edges and bipartite, 
# there are 4 degrees of freedom for these 8 cells
```

## Epimodel estimates
```{r, messages=messageCtrl, results=textRslt, collapse=TRUE}
# Dissolution model
coef.diss <- dissolution_coefs(dissolution = ~offset(edges), duration = 10)
coef.diss

# Sam's example (identifiable: 4 terms)
formation <- ~ edges + b1degree(0:1) + b2degree(0:1)
target.stats <- c(165, 100, 138, 135, 78)
est1 <- netest(nw, formation, target.stats, coef.diss)
```
```{r}
summary(est1)
round(est1$fit$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est1$fit)

# Add degrange (identifiable: 4 terms)
formation <- ~ edges + b1degree(0:1) + b2degree(0:1) + degrange(4)
target.stats <- c(165, 100, 138, 135, 78, 0)
est2 <- netest(nw, formation, target.stats, coef.diss)
```
```{r}
summary(est2)
round(est2$fit$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est2$fit)

# Add degree 2 for mode 1 only (overspecified: 5 terms)
formation <- ~ edges + b1degree(0:2) + b2degree(0:1) + degrange(4)
target.stats <- c(165, 100, 138, 10, 135, 78, 0)
est3 <- netest(nw, formation, target.stats, coef.diss)
```
```{r}
summary(est3)
round(est3$fit$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est3$fit)

# Add degree 2 for both modes (overspecified: 6 terms)
formation <- ~ edges + b1degree(0:2) + b2degree(0:2) + degrange(4)
target.stats <- c(165, 100, 138, 10, 135, 78, 25, 0)
est4 <- netest(nw, formation, target.stats, coef.diss)
```
```{r}
summary(est4)
round(est4$fit$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est4$fit)

# Add degree 3 (overspecified: 8 terms)
formation <- ~ edges + b1degree(0:3) + b2degree(0:3) + degrange(4)
target.stats <- c(165, 100, 138, 10, 2, 135, 78, 25, 12, 0)
est5 <- netest(nw, formation, target.stats, coef.diss)
```
```{r}
summary(est5)
round(est5$fit$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est5$fit)
```

## ERGM direct estimates

```{r, messages=messageCtrl, results=textRslt, collapse=TRUE}
# Sam's example (4 terms)
formation <- ~ edges + b1degree(0:1) + b2degree(0:1)
target.stats <- c(165, 100, 138, 135, 78)
est1.ergm <- ergm(nw ~ edges + b1degree(0:1) + b2degree(0:1),
            target.stats=target.stats)
```
```{r}
summary(est1.ergm)
round(est1.ergm$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est1.ergm)

# Add degrange (4 terms)
formation <- ~ edges + b1degree(0:1) + b2degree(0:1) + degrange(4)
target.stats <- c(165, 100, 138, 135, 78, 0)
est2.ergm <- ergm(nw~ edges + b1degree(0:1) + b2degree(0:1) + degrange(4),
                  target.stats=target.stats)
```
```{r}
summary(est2.ergm)
round(est2.ergm$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est2.ergm)

# Add degree 2 for mode 1 only (overspecified: 5 terms)
formation <- ~ edges + b1degree(0:2) + b2degree(0:1) + degrange(4)
target.stats <- c(165, 100, 138, 10, 135, 78, 0)
est3.ergm <- ergm(nw~ edges + b1degree(0:2) + b2degree(0:1) + degrange(4),
                    target.stats=target.stats)
```
```{r}
summary(est3.ergm)
round(est3.ergm$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est3.ergm)

# Add degree 2 for both modes (overspecified: 6 terms)
formation <- ~ edges + b1degree(0:2) + b2degree(0:2) + degrange(4)
target.stats <- c(165, 100, 138, 10, 135, 78, 25, 0)
est4.ergm <- ergm(nw~ edges + b1degree(0:2) + b2degree(0:2) + degrange(4),
                    target.stats=target.stats)
```
```{r}
summary(est4.ergm)
round(est4.ergm$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est4.ergm)

# Add degree 3 (overspecified: 8 terms)
formation <- ~ edges + b1degree(0:3) + b2degree(0:3) + degrange(4)
target.stats <- c(165, 100, 138, 10, 2, 135, 78, 25, 12, 0)
est5.ergm <- ergm(nw~ edges + b1degree(0:3) + b2degree(0:3) + degrange(4),
                    target.stats=target.stats)
```
```{r}
summary(est5.ergm)
round(est5.ergm$covar, 3)
```
```{r,messages=messageCtrl, results=textRslt, collapse=TRUE}
mcmc.diagnostics(est5.ergm)
```

## EpiModel-ERGM comparisons by model

Note: 

* the edges term will differ b/c EpiModel is adjusting for dissolution.  
* sometimes the overspecified models will converge on the same coef estimates, sometimes not.  this may be why the MCMC% is 100.   
* residual differences due to MCMC are quite small  
```{r}
cbind(EpiModel=round(est1$coef.form,4),
      ergm=round(est1.ergm$coef,4))
```

```{r}
cbind(EpiModel=round(est2$coef.form,4),
      ergm=round(est2.ergm$coef,4))
```

```{r}
# overspecified: 
cbind(EpiModel=round(est3$coef.form,4),
      ergm=round(est3.ergm$coef,4))
```

```{r}
# overspecified: 
cbind(EpiModel=round(est4$coef.form,4),
      ergm=round(est4.ergm$coef,4))
```

```{r}
# overspecified: 
cbind(EpiModel=round(est5$coef.form,4),
      ergm=round(est5.ergm$coef,4))
```
